---
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
---

```{r, include = FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(kableExtra)
library(igraph)
library(ggtext)
varorder <- c('LOGIMEM', 'MEMUNITS', 'DIGIF', 'DIGIB', 'WAIS', 'TRAILA', 'TRAILB', 'BOSTON', 'ANIMALS')
```

## Data Analysis

Using the MLLT model which allows correlation in the observation errors and underlying cognitive process (OS), we carry out an analysis on the NACC data with the outcomes and predictors of interest as described in (section__________). The Gibb's sampling was repeated 5,000 times with a burn-in of 2,000 samples. Non-informative priors are used for both the linear effects and covariance matrices.

### Data Analysis Results

#### Dementia

```{r}
SIG <- "Paper2/paperData/NACC_SIG_demen.RDS" %>%
  readRDS() %>%
  data.frame() %>%
  mutate_all(.funs = list(function(x)ifelse( x== 'red', 'green', ifelse(x == 'blue', 'red', x))))

ptab <- "Paper2/paperData/NACC_Effects_demen.RDS" %>%
  readRDS()%>%
  t() %>%
  data.frame()



ptab <- ptab[varorder]

for(i in 1:length(varorder)){
  ptab[i] <- cell_spec(ptab[[i]], background = SIG[[i]])
}


ptab[,1:5] %>%
  kbl(row.names = TRUE,  longtable = TRUE, escape = FALSE) %>%
  kable_styling(font_size = 7)
  

ptab[,6:9] %>%
  kbl(row.names = TRUE,  longtable = TRUE, escape = FALSE) %>%
    kable_styling(font_size = 7)
```


```{r}
readRDS("Paper2/paperData/NACC_Eps_demen.RDS") %>%
  mutate(V1 = factor(V1, levels = varorder), V2 = factor(V2, levels = varorder)) %>%
  ggplot(aes(x = V2, y = V1, fill = COR, label = print_values)) +
  xlab("") +
  ylab("") +
  theme_minimal()  +
  geom_tile( color = 'white')+
  geom_richtext(
        fill = NA, label.color = NA, # remove background and outline
        label.padding = grid::unit(rep(0, 4), "pt") # remove padding
  )  +
  scale_fill_gradient2(low = 'lightblue', high = "red", 
    midpoint = .5,limits = c(0,1), na.value = "lightblue",
    name="Correlation")+
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(angle = -30, vjust = 0, hjust=.35)
    # legend.justification = c(1, 0),
    # legend.position = c(0.4, 0.6),
    # legend.direction = "horizontal"
    ) +
  # guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
  #               title.position = "top", title.hjust = 0.5)) +
  ggtitle("Observation Correlation")
```



```{r}
readRDS("Paper2/paperData/NACC_Eta_demen.RDS") %>%
  mutate(V1 = factor(V1, levels = varorder), V2 = factor(V2, levels = varorder)) %>%
  ggplot(aes(x = V2, y = V1, fill = COR, label = print_values)) +
  xlab("") +
  ylab("") +
  theme_minimal()  +
  geom_tile( color = 'white')+
  geom_richtext(
        fill = NA, label.color = NA, # remove background and outline
        label.padding = grid::unit(rep(0, 4), "pt") # remove padding
  ) +
  scale_fill_gradient2(low = 'lightblue', high = "red", 
    midpoint = .5,limits = c(0,1), na.value = "lightblue",
    name="Correlation")+
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(angle = -30, vjust = 0, hjust=.35)
    # legend.justification = c(1, 0),
    # legend.position = c(0.4, 0.6),
    # legend.direction = "horizontal"
    ) +
  # guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
  #               title.position = "top", title.hjust = 0.5)) +
  ggtitle("State Correlation")
```






#### Non-Dementia

```{r}
SIG <- "Paper2/paperData/NACC_SIG_non_demen.RDS" %>%
  readRDS() %>%
  data.frame() %>%
  mutate_all(.funs = list(function(x)ifelse( x== 'red', 'green', ifelse(x == 'blue', 'red', x))))

ptab <- "Paper2/paperData/NACC_Effects_non_demen.RDS" %>%
  readRDS()%>%
  t() %>%
  data.frame()



ptab <- ptab[varorder]

for(i in 1:length(varorder)){
  ptab[i] <- cell_spec(ptab[[i]], background = SIG[[i]])
}


ptab[,1:5] %>%
  kbl(row.names = TRUE,  longtable = TRUE, escape = FALSE) %>%
  kable_styling(font_size = 7)
  

ptab[,6:9] %>%
  kbl(row.names = TRUE,  longtable = TRUE, escape = FALSE) %>%
    kable_styling(font_size = 7)
```


```{r}
readRDS("Paper2/paperData/NACC_Eps_non_demen.RDS") %>%
  mutate(V1 = factor(V1, levels = varorder), V2 = factor(V2, levels = varorder)) %>%
  ggplot(aes(x = V2, y = V1, fill = COR, label = print_values)) +
  xlab("") +
  ylab("") +
  theme_minimal()  +
  geom_tile( color = 'white')+
  geom_richtext(
        fill = NA, label.color = NA, # remove background and outline
        label.padding = grid::unit(rep(0, 4), "pt") # remove padding
  )  +
  scale_fill_gradient2(low = 'lightblue', high = "red", 
    midpoint = .5,limits = c(0,1), na.value = "lightblue",
    name="Correlation")+
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(angle = -30, vjust = 0, hjust=.35)
    # legend.justification = c(1, 0),
    # legend.position = c(0.4, 0.6),
    # legend.direction = "horizontal"
    ) +
  # guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
  #               title.position = "top", title.hjust = 0.5)) +
  ggtitle("Observation Correlation")
```



```{r}
readRDS("Paper2/paperData/NACC_Eta_non_demen.RDS") %>%
  mutate(V1 = factor(V1, levels = varorder), V2 = factor(V2, levels = varorder)) %>%
  ggplot(aes(x = V2, y = V1, fill = COR, label = print_values)) +
  xlab("") +
  ylab("") +
  theme_minimal()  +
  geom_tile( color = 'white')+
  geom_richtext(
        fill = NA, label.color = NA, # remove background and outline
        label.padding = grid::unit(rep(0, 4), "pt") # remove padding
  ) +
  scale_fill_gradient2(low = 'lightblue', high = "red", 
    midpoint = .5,limits = c(0,1), na.value = "lightblue",
    name="Correlation")+
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(angle = -30, vjust = 0, hjust=.35)
    # legend.justification = c(1, 0),
    # legend.position = c(0.4, 0.6),
    # legend.direction = "horizontal"
    ) +
  # guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
  #               title.position = "top", title.hjust = 0.5)) +
  ggtitle("State Correlation")
```


```{r}
fdo_demen <- readRDS("Paper2/paperData/FullDataOutput_demen.RDS")

fdo_non_demen <- readRDS("Paper2/paperData/FullDataOutput_non_demen.RDS")

demen_eps_cor <- map(2000:5000, function(i)cov2cor(fdo_demen$Eps.Track[,,i] )) %>%
  abind::abind(along = 3)

non_demen_eps_cor <- map(2000:5000, function(i)cov2cor(fdo_non_demen$Eps.Track[,,i] )) %>%
  abind::abind(along = 3)

lcl_eps_demen <- demen_eps_cor %>%
  apply(1:2, quantile, .025)



ucl_eps_demen <- demen_eps_cor %>%
  apply(1:2, quantile, .975)

lcl_eps_non_demen <- non_demen_eps_cor %>%
  apply(1:2, quantile, .025)

ucl_eps_non_demen <- non_demen_eps_cor %>%
  apply(1:2, quantile, .975)

ci_overlap <- (ucl_eps_demen > lcl_eps_non_demen & ucl_eps_demen < ucl_eps_non_demen)|(lcl_eps_demen > lcl_eps_non_demen & lcl_eps_demen < ucl_eps_non_demen)|(lcl_eps_demen < lcl_eps_non_demen & ucl_eps_demen > ucl_eps_non_demen)

colnames(ci_overlap) <- row.names(ci_overlap) <- colnames(lcl_eps_demen) <- row.names(lcl_eps_demen) <- colnames(ucl_eps_demen) <- row.names(ucl_eps_demen)  <- colnames(lcl_eps_non_demen) <- row.names(lcl_eps_non_demen) <- colnames(ucl_eps_non_demen) <- row.names(ucl_eps_non_demen) <- varorder

ci_overlap
```


```{r}
demen_eps_cor <- map(2000:5000, function(i)cov2cor(fdo_demen$Eps.Track[,,i] )) %>%
  abind::abind(along = 3)

non_demen_eps_cor <- map(2000:5000, function(i)cov2cor(fdo_non_demen$Eps.Track[,,i] )) %>%
  abind::abind(along = 3)
```

```{r}
all_eps_cor <- rbind(
  map(1:dim(demen_eps_cor)[3], function(i){
    x <- demen_eps_cor[,,i]
    row.names(x) <- colnames(x) <- varorder
    x %>%
      data.frame() %>%
      rownames_to_column("V1") %>%
      gather("V2", "COR", -V1) %>%
      filter(V1 != V2)
  }) %>%
    do.call("rbind", .) %>%
    mutate(dstatus = "demen"),
  map(1:dim(non_demen_eps_cor)[3], function(i){
    x <- non_demen_eps_cor[,,i]
    row.names(x) <- colnames(x) <- varorder
    x %>%
      data.frame() %>%
      rownames_to_column("V1") %>%
      gather("V2", "COR", -V1) %>%
      filter(V1 != V2)
  }) %>%
    do.call("rbind", .) %>%
    mutate(dstatus = "non_demen")
)

for(i in varorder){
  for(j in varorder){
    if(i != j){
      p <- all_eps_cor %>%
        filter(V1 == i, V2 == j) %>%
        ggplot(aes(x = COR, fill = dstatus)) +
        geom_density(alpha = .5) +
        ggtitle(paste(i, j, sep = " || "))
      print(p)
    }
    
  }
}
```


```{r}
eps_cor_ci_demen <- list(lcl_eps_demen, ucl_eps_demen, lcl_eps_non_demen, ucl_eps_non_demen) %>%
  map2(rep(c("LCL", "UCL"), 2), function(x, param){
    x %>%
      data.frame() %>%
      rownames_to_column("V1") %>%
      gather("V2", "COR", -V1) %>%
      filter(V1 != V2) %>%
      mutate(param = param)
  })

names(eps_cor_ci_demen) <- c("lcl_demen", "ucl_demen", "lcl_non_demen", "ucl_non_demen")
corbig_eps <- rbind(
  eps_cor_ci_demen[grepl("non_demen", names(eps_cor_ci_demen))] %>%
    do.call("rbind", .) %>%
    mutate(dstatus = "non_demen"),
  eps_cor_ci_demen[!grepl("non_demen", names(eps_cor_ci_demen))] %>%
    do.call("rbind", .) %>%
    mutate(dstatus = "demen")
) 
for(i in varorder){
  for(j in varorder){
    if(i != j){
      p <- corbig_eps %>%
        filter(V1 == i, V2 == j) %>%
        ggplot(aes(y = dstatus, x = COR, color = dstatus)) +
        geom_line(size = 4) + 
        geom_point(size = 6) +
        ggtitle(paste(i, j, sep = " || "))
      print(p)
    }
    
  }
}


```



```{r}
demen_eta_cor <- map(2000:5000, function(i)cov2cor(fdo_demen$Eta.Track[,,i] )) %>%
  abind::abind(along = 3)

non_demen_eta_cor <- map(2000:5000, function(i)cov2cor(fdo_non_demen$Eta.Track[,,i] )) %>%
  abind::abind(along = 3)

lcl_eta_demen <- demen_eta_cor %>%
  apply(1:2, quantile, .025)



ucl_eta_demen <- demen_eta_cor %>%
  apply(1:2, quantile, .975)

lcl_eta_non_demen <- non_demen_eta_cor %>%
  apply(1:2, quantile, .025)

ucl_eta_non_demen <- non_demen_eta_cor %>%
  apply(1:2, quantile, .975)

ci_overlap <- (ucl_eta_demen > lcl_eta_non_demen & ucl_eta_demen < ucl_eta_non_demen)|(lcl_eta_demen > lcl_eta_non_demen & lcl_eta_demen < ucl_eta_non_demen)|(lcl_eta_demen < lcl_eta_non_demen & ucl_eta_demen > ucl_eta_non_demen)

colnames(ci_overlap) <- row.names(ci_overlap) <- colnames(lcl_eta_demen) <- row.names(lcl_eta_demen) <- colnames(ucl_eta_demen) <- row.names(ucl_eta_demen)  <- colnames(lcl_eta_non_demen) <- row.names(lcl_eta_non_demen) <- colnames(ucl_eta_non_demen) <- row.names(ucl_eta_non_demen) <- varorder

ci_overlap
```



```{r}
all_eta_cor <- rbind(
  map(1:dim(demen_eta_cor)[3], function(i){
    x <- demen_eta_cor[,,i]
    row.names(x) <- colnames(x) <- varorder
    x %>%
      data.frame() %>%
      rownames_to_column("V1") %>%
      gather("V2", "COR", -V1) %>%
      filter(V1 != V2)
  }) %>%
    do.call("rbind", .) %>%
    mutate(dstatus = "demen"),
  map(1:dim(non_demen_eta_cor)[3], function(i){
    x <- non_demen_eta_cor[,,i]
    row.names(x) <- colnames(x) <- varorder
    x %>%
      data.frame() %>%
      rownames_to_column("V1") %>%
      gather("V2", "COR", -V1) %>%
      filter(V1 != V2)
  }) %>%
    do.call("rbind", .) %>%
    mutate(dstatus = "non_demen")
)

for(i in varorder){
  for(j in varorder){
    if(i != j){
      p <- all_eta_cor %>%
        filter(V1 == i, V2 == j) %>%
        ggplot(aes(x = COR, fill = dstatus)) +
        geom_density(alpha = .5) +
        ggtitle(paste(i, j, sep = " || "))
      print(p)
    }
    
  }
}
```


```{r}
eps_cor_ci_demen <- list(lcl_eta_demen, ucl_eta_demen, lcl_eta_non_demen, ucl_eta_non_demen) %>%
  map2(rep(c("LCL", "UCL"), 2), function(x, param){
    x %>%
      data.frame() %>%
      rownames_to_column("V1") %>%
      gather("V2", "COR", -V1) %>%
      filter(V1 != V2) %>%
      mutate(param = param)
  })

names(eps_cor_ci_demen) <- c("lcl_demen", "ucl_demen", "lcl_non_demen", "ucl_non_demen")
corbig_eps <- rbind(
  eps_cor_ci_demen[grepl("non_demen", names(eps_cor_ci_demen))] %>%
    do.call("rbind", .) %>%
    mutate(dstatus = "non_demen"),
  eps_cor_ci_demen[!grepl("non_demen", names(eps_cor_ci_demen))] %>%
    do.call("rbind", .) %>%
    mutate(dstatus = "demen")
) 
for(i in varorder){
  for(j in varorder){
    if(i != j){
      p <- corbig_eps %>%
        filter(V1 == i, V2 == j) %>%
        ggplot(aes(y = dstatus, x = COR, color = dstatus)) +
        geom_line(size = 4) + 
        geom_point(size = 6) +
        ggtitle(paste(i, j, sep = " || "))
      print(p)
    }
    
  }
}


```
