---
title: "Factor Justification"
output: html_document
---


```{r, eval = TRUE, include = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggtext)
knitr::opts_chunk$set(echo = FALSE)
ds <- "Paper2/paperData/NACC_stateCor.RDS" %>%
  readRDS()
varorder <- c('LOGIMEM', 'MEMUNITS', 'DIGIF', 'DIGIB', 'WAIS', 'TRAILA', 'TRAILB', 'BOSTON', 'ANIMALS', 'VEG')
```


```{r}
dseig <- eigen(ds)

facLoading <- t(t(dseig$vectors) * sqrt(dseig$values))
rownames(facLoading) <- row.names(ds)
colnames(facLoading) <- paste("PC_", 1:10, sep = "")

facLoading <- facLoading %>%
  data.frame() 
prop <- dseig$values / sum(dseig$values)


```

## Correlation Matrices from Real Analysis



```{r}
readRDS("Paper2/paperData/NACC_Eps.RDS") %>%
  mutate(V1 = factor(V1, levels = varorder), V2 = factor(V2, levels = varorder)) %>%
  ggplot(aes(x = V2, y = V1, fill = COR, label = print_values)) +
  xlab("") +
  ylab("") +
  theme_minimal()  +
  geom_tile( color = 'white')+
  geom_richtext(
        fill = NA, label.color = NA, # remove background and outline
        label.padding = grid::unit(rep(0, 4), "pt") # remove padding
  )  +
  scale_fill_gradient2(low = 'lightblue', high = "red", 
    midpoint = .5,limits = c(0,1), na.value = "lightblue",
    name="Correlation")+
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(angle = -30, vjust = 0, hjust=.35)
    # legend.justification = c(1, 0),
    # legend.position = c(0.4, 0.6),
    # legend.direction = "horizontal"
    ) +
  # guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
  #               title.position = "top", title.hjust = 0.5)) +
  ggtitle("Observation Correlation")
```


```{r}
readRDS("Paper2/paperData/NACC_Eta.RDS") %>%
  mutate(V1 = factor(V1, levels = varorder), V2 = factor(V2, levels = varorder)) %>%
  ggplot(aes(x = V2, y = V1, fill = COR, label = print_values)) +
  xlab("") +
  ylab("") +
  theme_minimal()  +
  geom_tile( color = 'white')+
  geom_richtext(
        fill = NA, label.color = NA, # remove background and outline
        label.padding = grid::unit(rep(0, 4), "pt") # remove padding
  ) +
  scale_fill_gradient2(low = 'lightblue', high = "red", 
    midpoint = .5,limits = c(0,1), na.value = "lightblue",
    name="Correlation")+
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_text(angle = -30, vjust = 0, hjust=.35)
    # legend.justification = c(1, 0),
    # legend.position = c(0.4, 0.6),
    # legend.direction = "horizontal"
    ) +
  # guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
  #               title.position = "top", title.hjust = 0.5)) +
  ggtitle("State Correlation")
```


## Computing Principal Components from State Correlation Matrix to Infer Factor Design

First, I take the eigenvalue decomposition from the correlation matrix. 

### How many factors should I choose?

To decide the number of factors, I take into account how many of the principle components it takes to reach 95% of the variance of the outcomes, which happens at about 4 factors.

```{r}
data.frame(
  pc = paste("PC_", 1:10, sep = ""),
  prop = round(prop, 3),
  cumprop = round(cumsum(prop), 3)
) %>%
  kbl() %>%
  kable_paper("hover", full_width = F)
```


### Factor Loadings

To get the factor loadings I use eigen vectors * sqrt(eigen values).

```{r}
facLoading %>%
  round(3) %>%
  select(PC_1:PC_4) %>%
  kbl() %>%
  kable_paper("hover", full_width = F)
```

I then do abs(loading)/sum(loading) for each column to see the relative loading of each test. This gets at which tests are most highly loaded onto the factor.

```{r}
facLoading %>%
  mutate_all(.funs = list(function(x)abs(x)/sum(abs(x)))) %>%
  round(3)%>%
  select(PC_1:PC_4) %>%
  kbl() %>%
  kable_paper("hover", full_width = F)
```


## Choosing form of G

To get the form of G, I look at the proportion of loading above some threshold. If it is above that threshold, we will call it a part of the given factor. A seemingly natural cut-off would be a loading of greater than 1/#outcomes = 1/10. This leads to the following form:


```{r}
facMat <- facLoading %>%
  mutate_all(.funs = list(function(x)abs(x)/sum(abs(x)))) %>%
  mutate_all(.funs = list(function(x)ifelse(x < 0.1, 0, 1))) %>%
  select(PC_1:PC_4) %>%
  as.matrix()

rbind(facMat, TOTAL = colSums(facMat)) %>%
  cbind(., TOTAL = rowSums(.))%>%
  kbl() %>%
  kable_paper("hover", full_width = F)

```

Every factor has 4 or 5 tests loaded onto it and each test loads on to 1-3 factors. This seems to be identifiable enough for our purposes.

## Next step

Compare this form of G to the form of G that each test is supposed to load as you showed me in that paper. 
