---
output:
  bookdown::pdf_document2
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{amsmath}
- \usepackage{algorithm,algorithmic}
---


```{r, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
```



## Fully Simulated Data Analysis

We simulated 100 subjects with between 2-13 observations from the following model:


\begin{equation*}
\begin{aligned}
\begin{bmatrix}
y_{ij1}\\
y_{ij2}\\
y_{ij3}
\end{bmatrix}
&= \begin{bmatrix}
\alpha_{ij1}\\
\alpha_{ij2}\\
\alpha_{ij3}
\end{bmatrix}
+ 
\begin{bmatrix}
\boldsymbol{x_{ij}\beta_1}\\
\boldsymbol{x_{ij}\beta_2}\\
\boldsymbol{x_{ij}\beta_3} 
\end{bmatrix} +
\begin{bmatrix}
\varepsilon_{ij1}\\
\varepsilon_{ij2}\\
\varepsilon_{ij3}
\end{bmatrix}, \ \ \ 
\begin{bmatrix}
\varepsilon_{ij1}\\
\varepsilon_{ij2}\\
\varepsilon_{ij3}
\end{bmatrix} 
\sim N(0, 
\begin{bmatrix}
15 & 2.4 & 1\\
2.4 & 15 & 1\\
1 & 1 & 10
\end{bmatrix}
)\\
\begin{bmatrix}
\alpha_{ij1}\\
\alpha_{ij2}\\
\alpha_{ij3}
\end{bmatrix} & = 
\begin{bmatrix}
\alpha_{i(j-1)1}\\
\alpha_{i(j-1)2}\\
\alpha_{i(j-1)3}
\end{bmatrix} +
\begin{bmatrix}
\eta_{ij1}\\
\eta_{ij2}\\
\eta_{ij3}
\end{bmatrix}, \ \ \
\begin{bmatrix}
\eta_{ij1}\\
\eta_{ij2}\\
\eta_{ij3}
\end{bmatrix} \sim N(0, \delta_{ij}
\begin{bmatrix}
5 & 3.7 & 0\\
3.7 & 5 & 0\\
0 & 0 & 2
\end{bmatrix}
)
\end{aligned}
\end{equation*}



## Fully Simulated Effect Coverage

```{r, fig.height=6}
fse <- readRDS("Paper2/JPrezTab/Fake_Effects.RDS") %>%
  map2(names(.), ~cbind(.x, method = .y) %>% mutate(beta = c(4, 2, 1, -3, 0, 1, 0, 0, 0))) %>%
  do.call("rbind", .) %>%
  mutate(
    Variable = case_when(
      Beta %in% paste("B", c(1, 4, 7), sep = "") ~ "X1",
      Beta %in% paste("B", c(1, 4, 7)+1, sep = "") ~ "X2",
      Beta %in% paste("B", c(1, 4, 7)+2, sep = "") ~ "X3"
    ), 
    Test = case_when(
      Beta %in% paste("B", 1:3, sep = "") ~ "Y1",
      Beta %in% paste("B", (1:3) + 3, sep = "") ~ "Y2",
      Beta %in% paste("B", (1:3) + 6, sep = "") ~ "Y3"
    ),
    BLCL = LCL - (AvgMean - avgBias),
    BUCL = UCL - (AvgMean - avgBias),
    Method = case_when(
      method == "Dub" ~ "Meas &\nLat",
      method == "St" ~ "Lat",
      method == "Obs" ~ "Meas",
      method == "Ind" ~ "Ind",
    ) %>%
      factor(levels = c("Ind", "Meas", "Lat", "Meas &\nLat")),
    Coverage = round(Coverage, 3)
  )

fse %>%
  select(Test, Variable, Beta = beta, Coverage, Method) %>%
  spread(Method, Coverage) %>%
  kbl(row.names = FALSE,  longtable = TRUE)
```

## Fully Simulated Linear Effect Bias

```{r, fig.height=6}
fse %>%
  ggplot(aes(x = Method, y = avgBias, color = Method, shape = Method)) +
  geom_hline(yintercept = 0) + 
  geom_point(size = 5) +
  geom_errorbar(aes(ymin = BLCL, ymax = BUCL)) +
  facet_grid(Test ~ Variable) +
  scale_color_manual(values = c("#405E92",   "#DF3116",     "#DFB916", "#409247"))  +
  scale_shape_manual(values = c(15,10,  9, 8))  +
  ylab("Avg Bias")  +
  theme_bw()
```



## Fully Simulated Correlated Measurement Error

```{r}
EpsL <- readRDS("Paper2/JPrezTab/Fake_Eps.RDS")

sigeps <- c(15, 15, 10)
SIGeps <- diag(sigeps)
SIGeps[1,2] <- SIGeps[2, 1] <- 2.4
SIGeps[1,3] <- SIGeps[3, 1] <- 1
SIGeps[2,3] <- SIGeps[3, 2] <- 1

EpsLdf <- EpsL %>%
  map("avg") %>%
  map(~.x[!lower.tri(.x)]) %>%
  do.call("cbind", .) %>%
  cbind(true = SIGeps[!lower.tri(SIGeps)]) %>%
  round(3) %>%
  data.frame

EpsLdf$param = (param = outer(1:3, 1:3, FUN = "paste", sep = ",")[!lower.tri(SIGeps)])

EpsLdf %>%
  select(param, true, ind, meas = obs, lat = st, `meas & lat` = dub) %>%
  kbl(row.names = FALSE,  longtable = TRUE)


```

## Fully Simulated Correlated Measurement Error Coverage

```{r}
EpsLdfcov <- EpsL %>%
  map("coverage") %>%
  map(~.x[!lower.tri(.x)]) %>%
  do.call("cbind", .) %>%
  round(3) %>%
  data.frame

EpsLdfcov$param = (param = outer(1:3, 1:3, FUN = "paste", sep = ",")[!lower.tri(SIGeps)])

EpsLdfcov %>%
  select(param, ind,  meas = obs, lat = st, `meas & lat` = dub) %>%
  kbl(row.names = FALSE,  longtable = TRUE)


```



```{r, eval = FALSE}
## Fully Simulated State Correlation Estimation
EpsLdfcor <- EpsL %>%
  map("avg") %>%
  map(cov2cor) %>%
  map(~.x[!lower.tri(.x)]) %>%
  do.call("cbind", .) %>%
  cbind(true = cov2cor(SIGeps)[!lower.tri(SIGeps)]) %>%
  round(3) %>%
  data.frame

EpsLdfcor$param = (param = outer(1:3, 1:3, FUN = "paste", sep = ",")[!lower.tri(SIGeps)])

EpsLdfcor %>%
  select(param, true, ind,  meas = obs, lat = st, `meas & lat` = dub) %>%
  filter(!(param %in% c("1,1", "2,2", "3,3"))) %>%
  kbl(row.names = FALSE,  longtable = TRUE)
```


## Fully Simulated Correlated Latent Variables

```{r}
Sigma <- c(5, 3.714281, 0, 3.714281, 5, 0, 0, 0, 2) %>%
  matrix(3, 3)

EtaL <- readRDS("Paper2/JPrezTab/Fake_Eta.RDS")



EtaLdf <- EtaL %>%
  map("avg") %>%
  map(~.x[!lower.tri(.x)]) %>%
  do.call("cbind", .) %>%
  cbind(true = Sigma[!lower.tri(Sigma)]) %>%
  round(3) %>%
  data.frame

EtaLdf$param = (param = outer(1:3, 1:3, FUN = "paste", sep = ",")[!lower.tri(Sigma)])

EtaLdf %>%
  select(param, true, ind,  meas = obs, lat = st, `meas & lat` = dub) %>%
  kbl(row.names = FALSE,  longtable = TRUE)


```





```{r, eval = FALSE}
## Fully Simulated State Correlation Estimation
EtaLdfcor <- EtaL %>%
  map("avg") %>%
  map(cov2cor) %>%
  map(~.x[!lower.tri(.x)]) %>%
  do.call("cbind", .) %>%
  cbind(true = cov2cor(Sigma)[!lower.tri(Sigma)]) %>%
  round(3) %>%
  data.frame

EtaLdfcor$param = (param = outer(1:3, 1:3, FUN = "paste", sep = ",")[!lower.tri(Sigma)])

EtaLdfcor %>%
  select(param, true, ind,  meas = obs, lat = st, `meas & lat` = dub) %>%
  filter(!(param %in% c("1,1", "2,2", "3,3"))) %>%
  kbl(row.names = FALSE,  longtable = TRUE)


```


## Fully Simulated State Variance Coverage

```{r}
EtaLdfcov <- EtaL %>%
  map("coverage") %>%
  map(~.x[!lower.tri(.x)]) %>%
  do.call("cbind", .) %>%
  round(3) %>%
  data.frame

EtaLdfcov$param = (param = outer(1:3, 1:3, FUN = "paste", sep = ",")[!lower.tri(SIGeps)])

EtaLdfcov %>%
  select(param, ind,  meas = obs, lat = st, `meas & lat` = dub) %>%
  kbl(row.names = FALSE,  longtable = TRUE)


```




## Full Simulated Standardized Effect Coverage

```{r}
seL <- readRDS("Paper2/JPrezTab/Fake_Stan.RDS")

seL %>%
  map("coverage")%>%
  map(~c(.x)) %>%
  do.call("cbind", .) %>%
  round(3) %>%
  data.frame %>%
  mutate(test = paste("Y", rep(1:3, 3), sep = ""), variable = paste("X", sort(rep(1:3, 3)), sep = "")) %>%
  select(test, variable, meas = obs, lat = state, `meas & lat` = dub) %>%
  arrange(test) %>%
  kbl(row.names = FALSE,  longtable = TRUE)
```


